---
title: "bile_acid_analysis"
output: html_document
date: '2024-07-18'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Maaslin2)
```

## Load meta

```{r load_meta}
bifido_responders = read_csv('/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/data/16S/16S_bifido_responder_status.csv') %>%
  mutate(participant_id = factor(participant_id))

final_participant_list = read.csv('/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/metadata/RAMP_final_participant_list.csv')

ramp_metadata_df = read.csv('/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/metadata/RAMP_participant_metadata.csv') %>%
  mutate(participant_id = factor(participant_id),
         final_grp = factor(final_grp, levels=c('Placebo', 'Low Dose 2FL', 'High Dose 2FL'))) %>%
  left_join(bifido_responders) %>%
  filter(participant_id %in% final_participant_list$participant_id)

```

## Load data

```{r load_data}

targeted_bile_results = read.csv('/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/data/BILE_ACIDS/240718_metabolon_targeted_bile_analysis.csv') %>%
  select(`Subject.or.Animal.ID`, `Time.Point`, Analyte, `Measured.Concentration..ng.mg.`, `Weight.Corrected.Concentration..ng.mg.`, `LLOQ..ng.mg.`, `ULOQ..ng.mg.`, `Dilution.Factor`, `Client.Sample.ID`) %>%
  rename(participant_id = 1,
         timepoint = 2,
         analyte = 3,
         measured_conc = 4, 
         corrected_conc = 5, 
         lloq = 6, 
         uloq = 7, 
         dilution_factor = 8,
         sample_id = 9) %>%
  select(sample_id, everything()) %>%
  mutate(participant_id = factor(participant_id)) %>%
  mutate(timepoint = ifelse(sample_id == 'RAMP91', 4, timepoint)) %>% 
  filter(timepoint %in% c(2, 3)) %>%
  mutate(timepoint = factor(timepoint)) %>% 
  rowwise() %>%
  mutate(measured_conc = case_when(
    measured_conc == 'BLOQ' ~ as.character(lloq),
    measured_conc == 'NR' ~ NA,
    TRUE ~ measured_conc
  ),
  corrected_conc = case_when(
    corrected_conc == 'BLOQ' ~ as.character(lloq),
    corrected_conc == 'NR' ~ NA,
    TRUE ~ corrected_conc
  )) %>%
  filter(!is.na(measured_conc)) %>%
  mutate(measured_conc = as.numeric(measured_conc),
         corrected_conc = as.numeric(corrected_conc)) %>%
  mutate(analyte = factor(analyte, levels=c('Cholic Acid', 'Chenodeoxycholic Acid', 'Glycocholic Acid', 'Glycochenodeoxycholic Acid', 'Taurocholic Acid', 'Taurochenodeoxycholic Acid', 'Deoxycholic Acid', 'Lithocholic Acid', 'Ursodeoxycholic Acid', 'Glycodeoxycholic Acid', 'Glycolithocholic Acid', 'Glycoursodeoxycholic Acid', 'Taurodeoxycholic Acid', 'Taurolithocholic Acid', 'Tauroursodeoxycholic Acid'))) %>% 
  mutate(timepoint = factor(ifelse(timepoint == 2, 'Week 0', 'Week 3'), levels=c('Week 0', 'Week 3')))

targeted_bile_results_with_meta = targeted_bile_results %>% 
  left_join(ramp_metadata_df, by='participant_id') %>% 
  group_by(analyte) %>%
  filter(corrected_conc < (mean(corrected_conc, na.rm=T)+3*sd(corrected_conc, na.rm=T))) %>%
  ungroup()

targeted_bile_results %>% 
  group_by(analyte) %>%
  summarise(mean(corrected_conc, na.rm=T),
            sd(corrected_conc, na.rm=T))
```

## Compositional analysis

```{r }

library(viridis)

bile_acid_rel_ab = targeted_bile_results_with_meta %>% 
  group_by(participant_id, timepoint) %>%
  arrange(participant_id, timepoint) %>%
  mutate(rel_ab = 100*corrected_conc / sum(corrected_conc, na.rm=T))

bile_acid_rel_ab %>% group_by(analyte) %>% summarise(tot = sum(corrected_conc)) %>% arrange(desc(tot))

low_bile = c('Glycoursodeoxycholic Acid', 'Taurolithocholic Acid', 'Glycolithocholic Acid', 'Tauroursodeoxycholic Acid', 'Taurodeoxycholic Acid')

bile_acid_rel_ab_collapsed = bile_acid_rel_ab %>% 
  mutate(analyte = as.character(analyte)) %>%
  mutate(analyte_alt = ifelse(analyte %in% low_bile, 'Other', analyte)) %>% 
  mutate(analyte_alt = factor(analyte_alt, levels = c('Cholic Acid', 'Chenodeoxycholic Acid', 'Glycocholic Acid', 'Glycochenodeoxycholic Acid', 'Taurocholic Acid', 'Taurochenodeoxycholic Acid', 'Deoxycholic Acid', 'Lithocholic Acid', 'Ursodeoxycholic Acid', 'Glycodeoxycholic Acid', 'Other'))) %>%
  mutate(final_grp = factor(final_grp, levels=c('Placebo', 'Low Dose 2FL', 'High Dose 2FL')))

ggplot(bile_acid_rel_ab_collapsed, aes(fill=factor(analyte_alt), y=corrected_conc, x=participant_id)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap(timepoint~final_grp, scales='free_x') +
  scale_fill_brewer(palette = 'Spectral', name = 'Analyte') +
  scale_y_continuous(limits = c(0, 5000), expand = c(0.0, 0)) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank()) +
  ylab('Concentration (ng/mg feces)') +
  xlab('Participand ID') 

ggplot(bile_acid_rel_ab_collapsed, aes(fill=factor(analyte_alt), y=rel_ab, x=participant_id)) +
  geom_bar(position='stack', stat='identity') +
  facet_wrap(timepoint~final_grp, scales='free_x') +
  scale_fill_brewer(palette = 'Spectral', name = 'Analyte') +
  scale_y_continuous(limits = c(0, 100), expand = c(0.0, 0)) +
  theme_classic() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank()) +
  ylab('Relative Abundance') +
  xlab('Participand ID') 

```


## By treatment group

#### Individual concentrations

```{r plot_bile}
all_bile_plot = ggplot(targeted_bile_results_with_meta, aes(x=timepoint, y=corrected_conc, fill=factor(final_grp))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Concentration (ng/mg feces)') +
  xlab('Timepoint') +
  facet_wrap(~analyte, scales='free_y', nrow=3) +
  scale_fill_brewer(palette='Set2') +
  scale_color_brewer(palette='Set2') +
  guides(fill=guide_legend(title="Treatment Group"))

ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/all_bile_acids.pdf',
       all_bile_plot,
       width=12, height=6)

wilcox.test(targeted_bile_results_with_meta %>% filter(analyte == 'Taurocholic Acid') %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 0') %>% pull(corrected_conc),
            targeted_bile_results_with_meta %>% filter(analyte == 'Taurocholic Acid') %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 3') %>% pull(corrected_conc))


```

#### Fold change

```{r fold_change}
bile_acid_fold_change = targeted_bile_results_with_meta %>% 
  select(participant_id, timepoint, analyte, measured_conc, lloq) %>%
  spread(key='timepoint', value='measured_conc') %>% 
  mutate(`Week 3` = ifelse(is.na(`Week 3`), as.numeric(lloq), `Week 3`),
         `Week 0` = ifelse(is.na(`Week 0`), as.numeric(lloq), `Week 0`)) %>%
  mutate(fc = `Week 3`/`Week 0`) %>% 
  left_join(ramp_metadata_df, by='participant_id') %>% 
  group_by(analyte) %>%
  filter(fc < (mean(fc, na.rm=T)+3*sd(fc, na.rm=T))) %>%
  ungroup() %>%
  mutate(log2_fc = log2(fc))

bile_acid_fold_change_plots = ggplot(bile_acid_fold_change, aes(x=final_grp, y=log2_fc, fill=factor(final_grp))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('log2(Fold-change)') +
  xlab('Timepoint') +
  facet_wrap(~analyte, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set2') +
  scale_color_brewer(palette='Set2') +
  guides(fill=guide_legend(title="Treatment Group"))


ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/bile_acid_fold_change.pdf',
       plot = bile_acid_fold_change_plots,
       width=16, height=6)


t.test(bile_acid_fold_change %>% filter(analyte == 'Lithocholic Acid') %>% filter(final_grp == 'High Dose 2FL') %>% pull(log2_fc), mu = 0, alternative = "two.sided")

t.test(bile_acid_fold_change %>% filter(analyte == 'Deoxycholic Acid') %>% filter(final_grp == 'High Dose 2FL') %>% pull(log2_fc), mu = 0, alternative = "two.sided")

wilcox.test(bile_acid_fold_change %>% filter(analyte == 'Taurodeoxycholic Acid') %>% filter(final_grp == 'High Dose 2FL') %>% pull(log2_fc),
       bile_acid_fold_change %>% filter(analyte == 'Taurodeoxycholic Acid') %>% filter(final_grp == 'Placebo') %>% pull(log2_fc))



```

#### Conjugation subset

```{r by_type}
bile_acid_annotations = targeted_bile_results_with_meta %>% 
  pull(analyte) %>% 
  unique() %>%
  as.data.frame() %>% 
  rename(analyte = 1) %>%
  mutate(conjugated = case_when(
    grepl('Glyco', analyte) ~ 'Conjugated',
    grepl('Tauro', analyte) ~ 'Conjugated',
    TRUE ~ 'Unconjugated'
  ),
  which_conjugated = case_when(
    grepl('Glyco', analyte) ~ 'Glycine-conjugated',
    grepl('Tauro', analyte) ~ 'Taurine-conjugated',
    TRUE ~ 'Unconjugated'
  ),
  primary_or_secondary = case_when(
    analyte == 'Cholic Acid' ~ 'Primary',
    analyte == 'Chenodeoxycholic Acid' ~ 'Primary',
    analyte == "Deoxycholic Acid" ~ 'Secondary',
    analyte == "Glycochenodeoxycholic Acid" ~ 'Primary',
    analyte == "Glycocholic Acid" ~ 'Primary',
    analyte == 'Glycodeoxycholic Acid' ~ 'Secondary',
    analyte == "Glycolithocholic Acid" ~ 'Secondary',
    analyte == 'Glycoursodeoxycholic Acid' ~ 'Secondary',
    analyte == 'Lithocholic Acid' ~ 'Secondary',
    analyte == 'Taurochenodeoxycholic Acid' ~ 'Primary',
    analyte == 'Taurocholic Acid' ~ 'Primary',
    analyte == 'Taurodeoxycholic Acid' ~ 'Secondary',
    analyte == 'Taurolithocholic Acid' ~ 'Secondary',
    analyte == 'Tauroursodeoxycholic Acid' ~ 'Secondary',
    analyte == 'Ursodeoxycholic Acid' ~ 'Secondary'
  ),
  primsec_conj = case_when(
    analyte == 'Cholic Acid' ~ 'Primary Unconjugated',
    analyte == 'Chenodeoxycholic Acid' ~ 'Primary Unconjugated',
    analyte == "Deoxycholic Acid" ~ 'Secondary Unconjugated',
    analyte == "Glycochenodeoxycholic Acid" ~ 'Primary Conjugated',
    analyte == "Glycocholic Acid" ~ 'Primary Conjugated',
    analyte == 'Glycodeoxycholic Acid' ~ 'Secondary Conjugated',
    analyte == "Glycolithocholic Acid" ~ 'Secondary Conjugated',
    analyte == 'Glycoursodeoxycholic Acid' ~ 'Secondary Conjugated',
    analyte == 'Lithocholic Acid' ~ 'Secondary Unconjugated',
    analyte == 'Taurochenodeoxycholic Acid' ~ 'Primary Conjugated',
    analyte == 'Taurocholic Acid' ~ 'Primary Conjugated',
    analyte == 'Taurodeoxycholic Acid' ~ 'Secondary Conjugated',
    analyte == 'Taurolithocholic Acid' ~ 'Secondary Conjugated',
    analyte == 'Tauroursodeoxycholic Acid' ~ 'Secondary Conjugated',
    analyte == 'Ursodeoxycholic Acid' ~ 'Secondary Unconjugated'
  ))

targeted_bile_results_with_meta_with_bile_annot = targeted_bile_results_with_meta %>% 
  left_join(bile_acid_annotations, by='analyte') 

bile_conjugation_summary = targeted_bile_results_with_meta_with_bile_annot %>% 
  group_by(participant_id, timepoint, primsec_conj) %>%
  summarise(total_conc = sum(measured_conc, na.rm = T)) %>% 
  left_join(ramp_metadata_df, by='participant_id')


bile_acid_categories_plot = ggplot(bile_conjugation_summary, aes(x=timepoint, y=total_conc, fill=factor(final_grp))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Concentration (ng/mg feces)') +
  xlab('Timepoint') +
  facet_wrap(~primsec_conj, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set2') +
  scale_color_brewer(palette='Set2') +
  guides(fill=guide_legend(title="Treatment Group"))


ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/bile_acid_categories.pdf',
       plot = bile_acid_categories_plot,
       width=8, height=6)

wilcox.test(bile_conjugation_summary %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 0') %>% filter(primsec_conj == 'Secondary Unconjugated') %>% pull(total_conc),
            bile_conjugation_summary %>% filter(final_grp == 'Placebo') %>% filter(timepoint == 'Week 0') %>% filter(primsec_conj == 'Secondary Unconjugated') %>% pull(total_conc))

wilcox.test(bile_conjugation_summary %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 0') %>% filter(primsec_conj == 'Secondary Unconjugated') %>% pull(total_conc),
            bile_conjugation_summary %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 3') %>% filter(primsec_conj == 'Secondary Unconjugated') %>% pull(total_conc))

```

#### Conjugation fold change


```{r }

log2fc_conjugaiton_summary = bile_conjugation_summary %>% 
  select(participant_id, timepoint, primsec_conj, total_conc) %>%
  spread(key='timepoint', value='total_conc') %>% 
  mutate(fc = `Week 3`/`Week 0`) %>% 
  left_join(ramp_metadata_df, by='participant_id') %>% 
  group_by(primsec_conj) %>%
  filter(fc < (mean(fc, na.rm=T)+3*sd(fc, na.rm=T))) %>%
  ungroup() %>%
  mutate(log2_fc = log2(fc))

bile_acid_category_fold_change = ggplot(log2fc_conjugaiton_summary, aes(x=final_grp, y=log2_fc, fill=factor(final_grp))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('log2(Fold-change)') +
  xlab('Timepoint') +
  facet_wrap(~primsec_conj, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set2') +
  scale_color_brewer(palette='Set2') +
  guides(fill=guide_legend(title="Treatment Group"))

ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/bile_acid_categories_fold_change.pdf',
       plot = bile_acid_category_fold_change,
       width=8, height=6)

wilcox.test(log2fc_conjugaiton_summary %>% filter(primsec_conj == 'Secondary Unconjugated') %>% filter(final_grp == 'High Dose 2FL') %>% pull(fc),
            log2fc_conjugaiton_summary %>% filter(primsec_conj == 'Secondary Unconjugated') %>% filter(final_grp == 'Placebo') %>% pull(fc))
```

#### Conjugation fraction

```{r }

conjugation_ratios = targeted_bile_results_with_meta_with_bile_annot %>% 
  group_by(participant_id, timepoint, conjugated) %>%
  summarise(total_conc = sum(measured_conc, na.rm = T)) %>% 
  left_join(ramp_metadata_df, by='participant_id') %>%
  spread(key='conjugated', value='total_conc') %>%
  mutate(ratio = Conjugated / Unconjugated) %>% 
  ungroup() %>%
  filter(ratio < mean(ratio)+3*sd(ratio))

bile_acid_conjugation_fraction_plot = ggplot(conjugation_ratios, aes(x=timepoint, y=ratio*100, fill=factor(final_grp))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Percent of bile acid pool that is conjugated (%)') +
  xlab('Timepoint') +
  scale_fill_brewer(palette='Set2') +
  scale_color_brewer(palette='Set2') +
  guides(fill=guide_legend(title="Treatment Group"))

ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/bile_acid_categories_fraction.pdf',
       plot = bile_acid_conjugation_fraction_plot,
       width=5, height=5)

wilcox.test(conjugation_ratios %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 0') %>% pull(ratio),
            conjugation_ratios %>% filter(final_grp == 'High Dose 2FL') %>% filter(timepoint == 'Week 3') %>% pull(ratio),
            paired=T)
```
#### Maaslin analysis

```{r }

problem_samples = c('RAMP94', 'RAMP104', 'RAMP24', 'RAMP86', 'RAMP102', 'RAMP92', 'RAMP4', 'RAMP50')

bile_abund_matrix_raw = targeted_bile_results_with_meta %>% 
  ungroup() %>%
  select(sample_id, analyte, measured_conc) %>%
  spread(key='analyte', value='measured_conc', fill=0.5) %>%
  column_to_rownames(var='sample_id')

bile_meta_df = targeted_bile_results_with_meta %>% 
  select(sample_id, participant_id, timepoint, final_grp, age, gender) %>%
  distinct() %>%
  mutate(final_grp = factor(final_grp, levels=c('Placebo', 'Low Dose 2FL', 'High Dose 2FL')),
    phase_grp = as.numeric(timepoint)*as.numeric(final_grp)) %>%
  filter(!sample_id %in% problem_samples) %>%
  column_to_rownames(var='sample_id')

bile_abund_matrix_filtered = bile_abund_matrix_raw[rownames(bile_meta_df), ]


all_groups_mas_bile_acid_grp <- Maaslin2(
  input_data = bile_abund_matrix_raw,
  input_metadata = bile_meta_df,
  output = "/home/mmcarter/user_data/Projects/RAMP/METABOLOMICS/MAASLIN",
  min_abundance = 0.0,
  min_prevalence = 0.0,
  normalization = "NONE",
  transform = "NONE",
  analysis_method = "LM",
  max_significance = 0.05,
  fixed_effects = c("final_grp", "phase_grp"),
  random_effects = c("participant_id"),
  reference = c("final_grp,Placebo"),
  correction = "BH",
  standardize = TRUE,
  cores = 1)


all_groups_mas_bile_acid_grp$results %>%
  filter(metadata == 'phase_grp') %>%
  arrange(pval)

```

##### Responders

```{r }

bile_abund_matrix_raw = targeted_bile_results_with_meta %>% 
  ungroup() %>%
  select(sample_id, analyte, measured_conc) %>%
  spread(key='analyte', value='measured_conc', fill=0.5) %>%
  column_to_rownames(var='sample_id')

bile_meta_resp_df = targeted_bile_results_with_meta %>% 
  select(sample_id, participant_id, timepoint, responder_status_delta, age, gender) %>%
  filter(!is.na(responder_status_delta)) %>%
  distinct() %>%
  mutate(responder_status_delta = factor(responder_status_delta, levels=c('nonresponder', 'responder')),
    phase_responder = as.numeric(timepoint)*as.numeric(responder_status_delta)) %>%
  column_to_rownames(var='sample_id')

bile_abund_matrix_filtered = bile_abund_matrix_raw[rownames(bile_meta_resp_df), ]


all_groups_mas_bile_acid_responder <- Maaslin2(
  input_data = bile_abund_matrix_filtered,
  input_metadata = bile_meta_resp_df,
  output = "/home/mmcarter/user_data/Projects/RAMP/METABOLOMICS/MAASLIN",
  min_abundance = 0.0,
  min_prevalence = 0.0,
  normalization = "TSS",
  transform = "LOG",
  analysis_method = "LM",
  max_significance = 0.05,
  fixed_effects = c("responder_status_delta", "phase_responder", "age", "gender"),
  random_effects = c("participant_id"),
  reference = c("final_grp,Placebo"),
  correction = "BH",
  standardize = TRUE,
  cores = 1)


all_groups_mas_bile_acid_responder$results %>%
  filter(metadata == 'phase_responder') %>%
  arrange(pval)
```



## By responder status

#### Individual concentration

```{r }
targeted_bile_results_with_meta_resp = targeted_bile_results_with_meta %>% 
  filter(!is.na(responder_status_delta))

all_bile_acids_resp_plot = ggplot(targeted_bile_results_with_meta_resp, aes(x=timepoint, y=corrected_conc, fill=factor(responder_status_delta))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Concentration (ng/mg feces)') +
  xlab('Timepoint') +
  facet_wrap(~analyte, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set1') +
  scale_color_brewer(palette='Set1') +
  guides(fill=guide_legend(title="Responder Status"))

ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/all_bile_acids_resp.pdf',
       all_bile_acids_resp_plot,
       width=18, height=7)

wilcox.test(targeted_bile_results_with_meta_resp %>% filter(analyte == 'Glycocholic Acid') %>% filter(responder_status_delta == 'responder') %>% filter(timepoint == 'Week 3') %>% pull(corrected_conc), 
            targeted_bile_results_with_meta_resp %>% filter(analyte == 'Glycocholic Acid') %>% filter(responder_status_delta == 'nonresponder') %>% filter(timepoint == 'Week 3') %>% pull(corrected_conc))
```

#### Fold change

```{r }
bile_acid_fold_change_resp = targeted_bile_results_with_meta_resp %>% 
  select(participant_id, timepoint, analyte, measured_conc, lloq) %>%
  spread(key='timepoint', value='measured_conc') %>% 
  mutate(`Week 3` = ifelse(is.na(`Week 3`), as.numeric(lloq), `Week 3`),
         `Week 0` = ifelse(is.na(`Week 0`), as.numeric(lloq), `Week 0`)) %>%
  mutate(fc = `Week 3`/`Week 0`) %>% 
  left_join(ramp_metadata_df, by='participant_id') %>% 
  group_by(analyte) %>%
  filter(fc < (mean(fc, na.rm=T)+3*sd(fc, na.rm=T))) %>%
  ungroup() %>%
  mutate(log2_fc = log2(fc))

bile_acid_fold_change_plots_resp = ggplot(bile_acid_fold_change_resp, aes(x=responder_status_delta, y=log2_fc, fill=factor(responder_status_delta))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('log2(Fold-change)') +
  xlab('Timepoint') +
  facet_wrap(~analyte, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set1') +
  scale_color_brewer(palette='Set1') +
  guides(fill=guide_legend(title="Responder Status"))


ggsave('/home/mmcarter/user_data/Projects/RAMP/BILE_ACIDS/figures/bile_acid_fold_change_responder.pdf',
       plot = bile_acid_fold_change_plots_resp,
       width=16, height=6)


t.test(bile_acid_fold_change_resp %>% filter(analyte == 'Glycodeoxycholic Acid') %>% filter(responder_status_delta == 'responder') %>% pull(log2_fc), mu = 0, alternative = "two.sided")

wilcox.test(bile_acid_fold_change_resp %>% filter(analyte == 'Glycochenodeoxycholic Acid') %>% filter(responder_status_delta == 'responder') %>% pull(log2_fc),
       bile_acid_fold_change_resp %>% filter(analyte == 'Glycochenodeoxycholic Acid') %>% filter(responder_status_delta == 'nonresponder') %>% pull(log2_fc),
       alternative='two.sided')
```

#### Conjugation Subset

```{r }
bile_conjugation_summary_resp = bile_conjugation_summary %>% 
  filter(!is.na(responder_status_delta))

ggplot(bile_conjugation_summary_resp, aes(x=timepoint, y=total_conc, fill=factor(responder_status_delta))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Concentration (ng/mg feces)') +
  xlab('Timepoint') +
  facet_wrap(~primsec_conj, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set1') +
  scale_color_brewer(palette='Set1') +
  guides(fill=guide_legend(title="Treatment Group"))

```

#### Conjugation fold change

```{r }
log2fc_conjugaiton_summary_resp = bile_conjugation_summary_resp %>% 
  select(participant_id, timepoint, primsec_conj, total_conc) %>%
  spread(key='timepoint', value='total_conc') %>% 
  mutate(fc = `Week 3`/`Week 0`) %>% 
  left_join(ramp_metadata_df, by='participant_id') %>% 
  group_by(primsec_conj) %>%
  filter(fc < (mean(fc, na.rm=T)+3*sd(fc, na.rm=T))) %>%
  ungroup() %>%
  mutate(log2_fc = log2(fc))

ggplot(log2fc_conjugaiton_summary_resp, aes(x=responder_status_delta, y=log2_fc, fill=factor(responder_status_delta))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('log2(Fold-change)') +
  xlab('Timepoint') +
  facet_wrap(~primsec_conj, scales='free_y', nrow=2) +
  scale_fill_brewer(palette='Set1') +
  scale_color_brewer(palette='Set1') +
  guides(fill=guide_legend(title="Treatment Group"))

wilcox.test(log2fc_conjugaiton_summary_resp %>% filter(responder_status_delta == 'responder') %>% filter(primsec_conj == 'Secondary Conjugated') %>% pull(log2_fc),
            log2fc_conjugaiton_summary_resp %>% filter(responder_status_delta == 'nonresponder') %>% filter(primsec_conj == 'Secondary Conjugated') %>% pull(log2_fc))
            
```


#### Conjugation fraction

```{r }
conjugation_ratios_resp = conjugation_ratios %>%
 filter(!is.na(responder_status_delta)) %>% 
  ungroup() %>%
  filter(ratio < mean(ratio)+3*sd(ratio))

ggplot(conjugation_ratios_resp, aes(x=timepoint, y=ratio*100, fill=factor(responder_status_delta))) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(pch=21, position=position_jitterdodge(), size=1.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=330, hjust=0),
        legend.position='bottom') +
  ylab('Percent of bile acid pool that is conjugated (%)') +
  xlab('Timepoint') +
  scale_fill_brewer(palette='Set1') +
  scale_color_brewer(palette='Set1') +
  guides(fill=guide_legend(title="Treatment Group"))


wilcox.test(conjugation_ratios_resp %>% filter(timepoint == 'Week 3') %>% filter(responder_status_delta == 'responder') %>% pull(ratio),
            conjugation_ratios_resp %>% filter(timepoint == 'Week 3') %>% filter(responder_status_delta == 'nonresponder') %>% pull(ratio))
```




## PCA 

```{r pca}
library(ggfortify)
library(MASS)
library(tidyverse)
df <- iris[1:4]
pca_res <- prcomp(df, scale. = TRUE)

autoplot(pca_res)

problem_samples = c('RAMP94', 'RAMP104', 'RAMP24', 'RAMP86', 'RAMP102', 'RAMP92', 'RAMP4', 'RAMP50')

bile_data_spread = targeted_bile_results_with_meta %>% 
  filter(timepoint == '3') %>%
  # filter(!sample_id %in% problem_samples) %>%
  select(sample_id, analyte, measured_conc, final_grp, responder_status_delta, bifido_abund_3, EDTA_plasma) %>% 
  #filter(!is.na(responder_status_delta)) %>%
  spread(key='analyte', value='measured_conc', fill=0.25) %>%
  column_to_rownames(var='sample_id') 
  
pca_res =  prcomp(bile_data_spread[5:19], scale. = TRUE)
autoplot(pca_res, data= bile_data_spread, colour = 'EDTA_plasma', loadings=T, loadings.label=T, loadings.label.size=2)

ggplot(bile_data_spread, aes(x=EDTA_plasma, y=`Tauroursodeoxycholic Acid`)) +
  geom_point() +
  stat_smooth(method='lm')

summary(lm(`Tauroursodeoxycholic Acid` ~ EDTA_plasma, data=bile_data_spread))
```


## Microbiome correlations

### Week 3 comparions 

```{r }
ramp_ps_trim_genus_glom = readRDS('/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/data/16S/phyloseq_object_trimmed_genus_glom.rds')

# Assuming `phyloseq_object` is your phyloseq object and `bile_acid_df` is your bile acid data frame
# Extract the OTU (taxa) abundance data from the phyloseq object
metadata_df <- as.data.frame(as(sample_data(ramp_ps_trim_genus_glom), "matrix")) %>% 
  rownames_to_column(var='sample_id') %>% 
  select(sample_id, participant_id, timepoint)

# Merge metadata with microbial abundance and bile acid data
microbial_abundance_df <- as.data.frame(as(otu_table(ramp_ps_trim_genus_glom), "matrix")) %>%
  rownames_to_column("sample_id") %>%
  left_join(metadata_df, by = "sample_id") %>%
  filter(timepoint == '3') %>%
  select(-c(sample_id, timepoint)) %>%
  column_to_rownames(var='participant_id')

bile_acid_df = targeted_bile_results_with_meta %>% 
  filter(timepoint == 'Week 3') %>%
  select(participant_id, analyte, measured_conc) %>%
  spread(key='analyte', value='measured_conc') %>% 
  column_to_rownames(var='participant_id')

# Ensure that samples in both datasets match by sample names
common_samples <- intersect(rownames(microbial_abundance_df), rownames(bile_acid_df))
microbial_abundance_df <- microbial_abundance_df[common_samples, ]
bile_acid_df <- bile_acid_df[common_samples, ]

# Initialize matrices to store correlation coefficients and p-values
cor_matrix <- matrix(NA, nrow = ncol(microbial_abundance_df), ncol = ncol(bile_acid_df))
p_matrix <- matrix(NA, nrow = ncol(microbial_abundance_df), ncol = ncol(bile_acid_df))

# Perform pairwise correlation and significance testing
for (i in 1:ncol(microbial_abundance_df)) {
  for (j in 1:ncol(bile_acid_df)) {
    cor_test <- cor.test(microbial_abundance_df[, i], bile_acid_df[, j], method = "pearson")
    cor_matrix[i, j] <- cor_test$estimate
    p_matrix[i, j] <- cor_test$p.value
  }
}

# Adjust p-values using the Benjamini-Hochberg method
p_adjusted <- p.adjust(p_matrix, method = "BH")

p_adjusted_matrix = matrix(p_adjusted, nrow = ncol(microbial_abundance_df), ncol = ncol(bile_acid_df))

# Set row and column names of the matrices for clarity
rownames(cor_matrix) <- colnames(microbial_abundance_df)
colnames(cor_matrix) <- colnames(bile_acid_df)
rownames(p_matrix) <- colnames(microbial_abundance_df)
colnames(p_matrix) <- colnames(bile_acid_df)
rownames(p_adjusted_matrix) <- colnames(microbial_abundance_df)
colnames(p_adjusted_matrix) <- colnames(bile_acid_df)

# Create a logical matrix indicating statistically significant correlations (e.g., p < 0.05)
significant_matrix <- matrix(p_adjusted < 0.05, nrow = ncol(microbial_abundance_df), ncol = ncol(bile_acid_df))
rownames(significant_matrix) <- colnames(microbial_abundance_df)
colnames(significant_matrix) <- colnames(bile_acid_df)

border_matrix <- matrix(NA, nrow = nrow(significant_matrix), ncol = ncol(significant_matrix))

# Set border color for significant cells (e.g., black border)
border_matrix[significant_matrix] <- "green"

library(pheatmap)
library(viridis)
# Heatmap of correlation coefficients
pheatmap(cor_matrix, cluster_rows = T, cluster_cols = T, 
         border_color = border_matrix,
         cellheight=5, cellwidth = 15,
         display_numbers = significant_matrix, # Display significance on heatmap
         fontsize = 4,
         main = "Correlation Heatmap of Microbial Taxa and Bile Acids",
         color =colorRampPalette(c("blue", "white", "red"))(50))

tax_table_otu = as.data.frame(as(tax_table(ramp_ps_trim_genus_glom), "matrix")) %>%
  rownames_to_column(var='OTU')

# Cluster rows (microbial taxa)
row_clustering <- hclust(dist(cor_matrix))
row_order <- row_clustering$order  # Get the order of rows

# Cluster columns (bile acids)
col_clustering <- hclust(dist(t(cor_matrix)))
col_order <- col_clustering$order  # Get the order of columns

# Convert matrices to a data frame in long format
cor_df <- as.data.frame(as.table(cor_matrix))  # Long format for correlations
p_df <- as.data.frame(as.table(p_adjusted_matrix))    # Long format for p-values


# Merge the correlation and p-value data frames
plot_df <- cor_df %>%
  rename(Microbial_Taxon = Var1, Bile_Acid = Var2, Correlation = Freq) %>%
  left_join(p_df %>% rename(Microbial_Taxon = Var1, Bile_Acid = Var2, P_Value = Freq), 
            by = c("Microbial_Taxon", "Bile_Acid")) %>%
  left_join(tax_table_otu, by=c('Microbial_Taxon'='OTU'))





# Step 3: Reorder rows and columns based on clustering
# Convert the factor levels of Microbial_Taxon and Bile_Acid to reflect the cluster ordering
plot_df$Microbial_Taxon <- factor(plot_df$Microbial_Taxon, levels = rownames(cor_matrix)[row_order])
plot_df$Bile_Acid <- factor(plot_df$Bile_Acid, levels = colnames(cor_matrix)[col_order])

taxon_order <- plot_df %>%
  arrange(Microbial_Taxon) %>%  # Ensure `Microbial_Taxon` is ordered by clustering
  select(Microbial_Taxon, Genus) %>%
  distinct(Genus, .keep_all = TRUE)

plot_df$Genus <- factor(plot_df$Genus, levels = taxon_order$Genus)

# Create the bubble plot using ggplot2
bile_otu_bubble_plot = ggplot(plot_df, aes(x = Bile_Acid, y = Genus, size = -log10(P_Value), color = Correlation)) +
  geom_point() +
  scale_size_continuous(range = c(1, 5), name = "Significance (-log10 P)") +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Correlation") +
  theme_minimal() +
  labs(x = "Bile Acid",
       y = "Microbial Taxon") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
        axis.text.y = element_text(size=5))  # Rotate x-axis labels for better readability

ggsave(filename = '/home/mmcarter/user_data/Projects/RAMP/RAMP_study_analysis/plots/bile_otu_bubble_plot.pdf',
       plot = bile_otu_bubble_plot,
       width=6, height=5)

p_df %>% arrange(Freq) %>%
  left_join(tax_table_otu, by=c('Var1'='OTU')) %>%
  select(Genus, Var2, Freq) %>%
  filter(Freq < 0.05)
```

```{r }
significant_pairs <- which(significant_matrix, arr.ind = TRUE)
# Create a list to store the plots
plots_list <- list()

# Loop through each significant pair and create scatterplots
for (k in 1:nrow(significant_pairs)) {
  # Get the row (microbial taxon) and column (bile acid) indices
  microbial_index <- significant_pairs[k, 1]
  bile_acid_index <- significant_pairs[k, 2]
  
  # Get the microbial taxon and bile acid names
  microbial_taxon <- colnames(microbial_abundance_df)[microbial_index]
  bile_acid <- colnames(bile_acid_df)[bile_acid_index]
  
  # Prepare the raw data for plotting
  raw_data <- data.frame(
    Microbial_Abund = microbial_abundance_df[, microbial_index],
    Bile_Acid_Abund = bile_acid_df[, bile_acid_index]
  )
  
  # Create the scatterplot
  plot <- ggplot(raw_data, aes(x = Microbial_Abund, y = Bile_Acid_Abund)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste(microbial_taxon, "vs", bile_acid),
         x = paste("Fold Change in", microbial_taxon),
         y = paste("Fold Change in", bile_acid)) +
    theme_minimal()
  
  # Add the plot to the list
  plots_list[[k]] <- plot
}

# Display all the plots
# You can use gridExtra or patchwork to arrange multiple ggplot plots if needed
library(gridExtra)
do.call(grid.arrange, c(plots_list, ncol = 2))  # Adjust ncol as needed
```


### Fold-change comparisons
```{r }
# Load necessary libraries
library(phyloseq)
library(tidyverse)
library(pheatmap)
library(compositions)  # For CLR transformation

# Assuming `phyloseq_object` is your phyloseq object and `bile_acid_df` is your bile acid data frame
# Extract the OTU (taxa) abundance data from the phyloseq object
microbial_abundance <- as(otu_table(ramp_ps_trim_genus_glom), "matrix")

# You must have a metadata file that contains the information about individuals and sample times (baseline and end)
# Assuming `metadata_df` contains columns: "SampleID", "PersonID", "Timepoint" (where Timepoint is 'baseline' or 'end')
# Make sure metadata matches the sample IDs
metadata_df <- as(sample_data(ramp_ps_trim_genus_glom), 'matrix') %>% 
  as.data.frame() %>%
  rownames_to_column(var='SampleID') %>%
  # filter(treatment_grp == '91') %>%
  select(SampleID, participant_id, timepoint)

# Merge metadata with microbial abundance and bile acid data
microbial_abundance_fc_df <- as.data.frame(microbial_abundance) %>%
  rownames_to_column("SampleID") %>%
  left_join(metadata_df, by = "SampleID") %>%
  select(-c(SampleID)) %>%
  gather(key='OTU', value='count', -participant_id, -timepoint) %>%
  mutate(count = log10(count+1)) %>%
  filter(timepoint %in% c('2', '3')) %>%
  spread(key='timepoint', value='count', fill=0.) %>%
  mutate(fold_change = log2((`3`+1)/(`2`+1))) %>%
  select(participant_id, OTU, fold_change) %>%
  spread(key='OTU', value='fold_change') %>%
  column_to_rownames(var='participant_id') %>%
  scale()

bile_acid_fc_df = targeted_bile_results_with_meta %>% 
  select(participant_id, timepoint, analyte, corrected_conc) %>%
  spread(key='timepoint', value='corrected_conc', fill=0.25) %>% 
  mutate(fold_change = log2(`Week 3`/`Week 0`)) %>%
  # filter(analyte == 'Tauroursodeoxycholic Acid')
  select(participant_id, analyte, fold_change) %>%
  spread(key='analyte', value='fold_change', fill=0.) %>%
  column_to_rownames(var='participant_id') %>%
  scale()

# Ensure that samples in both datasets match by sample names
common_samples <- intersect(rownames(microbial_abundance_fc_df), rownames(bile_acid_fc_df))
microbial_abundance_fc_df <- microbial_abundance_fc_df[common_samples, ]
bile_acid_fc_df <- bile_acid_fc_df[common_samples, ]

# Initialize matrices to store correlation coefficients and p-values
cor_matrix <- matrix(NA, nrow = ncol(microbial_abundance_fc_df), ncol = ncol(bile_acid_fc_df))
p_matrix <- matrix(NA, nrow = ncol(microbial_abundance_fc_df), ncol = ncol(bile_acid_fc_df))

# Perform pairwise correlation and significance testing
for (i in 1:ncol(microbial_abundance_fc_df)) {
  for (j in 1:ncol(bile_acid_fc_df)) {
    cor_test <- cor.test(microbial_abundance_fc_df[, i], bile_acid_fc_df[, j], method = "pearson")
    cor_matrix[i, j] <- cor_test$estimate
    p_matrix[i, j] <- cor_test$p.value
  }
}

# Adjust p-values using the Benjamini-Hochberg method
p_adjusted <- p.adjust(p_matrix, method = "none")

# Set row and column names of the matrices for clarity
rownames(cor_matrix) <- colnames(microbial_abundance_fc_df)
colnames(cor_matrix) <- colnames(bile_acid_fc_df)
rownames(p_matrix) <- colnames(microbial_abundance_fc_df)
colnames(p_matrix) <- colnames(bile_acid_fc_df)

# Create a logical matrix indicating statistically significant correlations (e.g., p < 0.05)
significant_matrix <- matrix(p_adjusted < 0.05, nrow = ncol(microbial_abundance_fc_df), ncol = ncol(bile_acid_fc_df))
rownames(significant_matrix) <- colnames(microbial_abundance_fc_df)
colnames(significant_matrix) <- colnames(bile_acid_fc_df)

border_matrix <- matrix(NA, nrow = nrow(significant_matrix), ncol = ncol(significant_matrix))

# Set border color for significant cells (e.g., black border)
border_matrix[significant_matrix] <- "green"

# Heatmap of correlation coefficients
pheatmap(cor_matrix, cluster_rows = TRUE, cluster_cols = TRUE, 
         cellheight=5, cellwidth = 15,
         border_color = border_matrix,
         display_numbers = significant_matrix, # Display significance on heatmap
         fontsize = 4,
         main = "Correlation Heatmap of Microbial Taxa and Bile Acids",
         color =colorRampPalette(c("blue", "white", "red"))(50))
```

```{r }
significant_pairs <- which(significant_matrix, arr.ind = TRUE)[1:5, ]
# Create a list to store the plots
plots_list <- list()

# Loop through each significant pair and create scatterplots
for (k in 1:nrow(significant_pairs)) {
  # Get the row (microbial taxon) and column (bile acid) indices
  microbial_index <- significant_pairs[k, 1]
  bile_acid_index <- significant_pairs[k, 2]
  
  # Get the microbial taxon and bile acid names
  microbial_taxon <- colnames(microbial_abundance_fc_df)[microbial_index]
  bile_acid <- colnames(bile_acid_fc_df)[bile_acid_index]
  
  # Prepare the raw data for plotting
  raw_data <- data.frame(
    Microbial_Abund = microbial_abundance_fc_df[, microbial_index],
    Bile_Acid_Abund = bile_acid_fc_df[, bile_acid_index]
  )
  
  # Create the scatterplot
  plot <- ggplot(raw_data, aes(x = Microbial_Abund, y = Bile_Acid_Abund)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste(microbial_taxon, "vs", bile_acid),
         x = paste("Fold Change in", microbial_taxon),
         y = paste("Fold Change in", bile_acid)) +
    theme_minimal()
  
  # Add the plot to the list
  plots_list[[k]] <- plot
}

# Display all the plots
# You can use gridExtra or patchwork to arrange multiple ggplot plots if needed
library(gridExtra)
do.call(grid.arrange, c(plots_list, ncol = 2))  # Adjust ncol as needed
```

